How the Lattice Boltzmann Method changed my life forever!

### 2016-07-13 thesis.tex
minor edits

% MG: replace \ \cite with ~\cite to ensure citations stay on same line
% MG: need an overall consistency in 'flucuation' vs. 'oscillation' vs. 'noise'; use same word if you mean same thing.
% MG: same note as above for 'local equilibrium' vs. 'quasiequilibrium'
% MG: consider using collision frequency, \omega, instead of relaxation time, \tau, so that the relaxation time is not confused with shear stress

@@

\title{How the Lattice Boltzmann Method changed my life forever!}
% The optional argument is the version of the title that will appear in Acrobat Reader's Document Info dialog box.
\author{Matthew Grasinger}
\degree{BS in Civil and Environmental Engineering, University of Pittsburgh, Pittsburgh, PA 2013}
\date{July 20, 2016}
\keywords{lattice Boltzmann method, non-Newtonian flow, free surface flow, wellbore cementing}
\subject{Draft}
\school{Swanson School of Engineering}
\degreesought{Master of Science}

@@

Despite the advantages of LBM, there are considerations that must be made, as with all computational methods, in regards to the accuracy and numerical stability of the solution.
Issues with accuracy and numerical stability are especially prevalent in non-Newtonian flows because of the nonlinear constitutive relationship.
\Fref{cha:lbm-stability} is a numerical investigation of the accuracy, stability, and computational efficiency of different LBM models in simulating non-Newtonian flows.

Once the ground-work has been laid for simulating non-Newtonian flows using LBM, the algorithm was extended in \Fref{cha:lbm-cement} to incorporate simulation of free-surface flows.
The extended LBM model was used to study primary cementing of a dry annulus, i.e. an annulus that is not filled with drilling mud.
More specifically, the study involved parameterizing different cement slurry flows and investigating how well each slurry flow filled different wellbore geometries.
The study is followed by conclusions and a discussion of future work.

@@

\chapter{Numerical Investigation of the Accuracy, Stability, and Efficiency of Lattice Boltzmann Models in Simulating Non-Newtonian Flows} \label{cha:lbm-stability}

@@

The dynamic viscosity of a fluid is a measure of its resistance to shear deformation.
For many fluids, at a constant temperature, the dynamic viscosity can be considered as constant.
These fluids are known as Newtonian fluids.
Non-Newtonian fluids have an apparent dynamic viscosity, or apparent resistance to shear stress, that is variable--even at a constant temperature--and is often a function of strain-rate.
There are a number of fluids in science and engineering applications that can be classified as non-Newtonian; pastes, slurries, molten plastics, polymer solutions, dyes, varnishes, suspensions, and some biomedical liquids such as blood all behave in a non-Newtonian manner \cite{bohme1987non}.
Of all of the different non-Newtonian behaviors that exist, there are two models under which much of the behaviors may be idealized:
\begin{enumerate}
	\item \emph{Yield stress fluids}. Yield stress fluids, also known as Bingham plastics, do not flow until a threshold value of stress, referred to as its yield stress, is exceeded.
	Some examples of yield stress fluids are drilling muds, fresh cement slurries and concretes, and melted metal alloys \cite{ginzburg2002free}.
	Understanding the flow of cement slurries and concretes is important because they are used in the cementing of oil and gas wellbores, and in construction.
	If the cement job for an oil and gas wellbore is not completed successfully then gases and fluids can leak through the wellbore to contaminate ground water or the atmosphere~\cite{grasinger2015simulation}. %MG: is it tacky to cite myself here?
	The goal of understanding cement slurry and concrete flows during construction is to understand how construction forms can be completely filled as voids can lead to structural deficiencies~\cite{definitely cite someone}.
	Yield stress flow is relevant in many other disciplines and applications because of the many substances that exhibit yield stress behavior, e.g. pastes, paints, muds, molten plastics and metals, and blood \cite{wang2011lattice}. %MG: citation is for blood as non-Newtonian
	As a consequence, simulating yield stress fluids can help engineers develop better pastes and paints, understand the flow and deformation of mud and clay in geotechnical engineering, understand blood circulation, and better manufacture metal and plastic parts.

@@

Analytical solutions rarely exists for even the simplest non-Newtonian flows because of the complexity that a nonlinear constitutive relationship entails.
It is generally more practical to approximate non-Newtonian flows using numerical methods.
However, the nonlinear constitutive equation--typically of the form \begin{equation} \label{eq:non-newtonian}
\tau = \mu_{app}(\dot{\gamma}) \dot{\gamma},
\end{equation}\noindent where the apparent viscosity, $\mu_{app}$ is a function of the strain rate--results in certain challenges for numerical methods as well.
%Constitutive form for generalized Newtonian fluids, $\tau = \mu(\dot{\gamma}) \dot{\gamma}$, where the apparent viscosity, $\mu$ is a function of the strain rate
Determining the apparent viscosity and shear rate of a flow will often require an iterative solution, a general Picard-type algorithm is:

@@

One such consideration associated with this view, is the potential for numerical inaccuracies and instabilities ~\cite{sterling1993stability,sterling1996stability,bawazeer2013stability,lallemand2000theory}. %MG: are all relevant sources cited?

@@

Much work has been done (not specific to non-Newtonian flows) to investigate and develop ways to enhance the stability of lattice Boltzmann methods.
\citet{sterling1993stability,sterling1996stability} investigated the stability of LBM with the linear Bhatnagar--Gross--Krook (BGK) collision operator on 7-velocity and 9-velocity 2D lattices and a 15-velocity cubic lattice.
The investigation was performed by linearizing all nonlinear terms about global equilibrium values and applying a von Neumann stability analysis on the resulting linearized equations.
The spatial dependence of stability was analyzed by taking the Fourier transform of the linearized equations.
\citeauthor{sterling1993stability} concluded that the linear stability of LBM models depends on the mass distribution parameters, the mean velocity, the relaxation time, and the wave number.
~\citet{worthing1997stability} extended the stability analysis performed by \citeauthor{sterling1993stability} and discovered both physical and nonphysical instabilities; and that setting the BGK collision frequency, $\omega = 1$ provides the optimal accuracy in time.

@@

To improve upon the stability of LBM models, \citet{d1994generalized} developed a multiple-relaxation-time (MRT) collision operator, that takes place in moment space and allows each moment to relax at a different rate.
\citet{lallemand2000theory} used von Neumann stability analysis to investigate the stability of the newly constructed LBM-MRT model and concluded LBM with the MRT collision operator was more stable, but with increased computational expense, than what is still often used, the BGK collision operator.
Note that although this increased computational expense was decided to be not too significant for Newtonian flows ($\approx$ 10-20\% \cite{lallemand2000theory}), the issue may be magnified for non-Newtonian flows because an iterative solution for the constitutive equation can require that certain expensive computations be performed at each iteration (the strain rate tensor is determined from the nonequilibrium distribution function which must be mapped into momentum space when using the MRT collision operator).
Both the BGK~\cite{wang2011lattice,ashrafizaadeh2009comparison,wang2015localized,tang2011bingham} and MRT~\cite{chen2014simulations,fallah2012multiple,chai2011multiple} collision operators have been used successfully to simulate non-Newtonian flows.
\citeauthor{chen2014simulations} concluded that the MRT collision operator was more stable for Bingham plastic flow and allowed the use of a more accurate approximation to the Bingham plastic constitutive relationship.
However, what remains unclear is:
\begin{itemize}
    \item What is the increased cost associated with the MRT collision operator when applied to non-Newtonian flows?
    \item Under what conditions, e.g. material parameters, physical problem, etc., for Bingham plastic fluid flows is the MRT collision operator necessary to maintain stability and/or accuracy?
    \item Under what conditions, e.g. material parameters, physical problem, etc, for power-law fluid flows is the MRT collision operator necessary to maintain stability and/or accuracy?
    \item What are additional strategies for increasing stability and accuracy, and what are their associated computational costs?
\end{itemize}
In regards to the last question, much work has been done recently to enhance stability of LBM models beyond the MRT collision operator.
\citet{brownlee2006stabilization,brownlee2007stability,brownlee2008nonequilibrium,packwood2009entropy,gorban2014enhancement} have all developed and tested means for introducing artificial dissipation in order to dampen out high frequency, nonphysical oscillations--oscillations that often lead to numerical instabilities if left unchecked.
Stability enhancement through artificial dissipation and entropic filtering has shown a lot of promise, but to the authors' knowledge has not been tested for use in non-Newtonian flows.

@@

\subsection{The Boltzmann Equation}
The Boltzmann equation (BE) can be thought of as a conservation of particle distributions.
The BE is given as $\frac{\partial f}{\partial t} + \pvel \cdot \grad f = \colop$, where $f = f(\pos, \pvel, t)$ is the particle velocity distribution function, $\pos$ is the spatial position vector, $\pvel$ is the particle velocity, $\mvel$ is the macroscopic velocity, and $\colop$ is the collision operator.
The lattice part of LBM refers to the way in which the BE is discretized.
The lattice discretizes the spatial domain with nodes that are connected to their neighbors through discrete lattice velocity vectors.
The velocity vectors act as pathways for particle distributions to travel along.
Each time step in LBM consists of two distinct actions:

\begin{itemize}
\item Streaming: particle distribution propagate to their neighbors along the lattice velocity vectors.
The particles can only move along the vectors in their specified direction and can only move at a specific speed.
\item Collision: particle distributions meet at a node and ``collide''.
In LBM, collisions are not simulated in a realistic sense, meaning that each individual particle does not exist and glance off of, or interact with, one another.
Instead the collision operator is formulated in such a way that particle distributions are relaxed toward equilibrium.
What defines equilibrium depends on the mechanics of interest to be modeled.
\end{itemize}

@@

\begin{figure}
  \centering
  \includegraphics[]{figs/d2q9}
  \caption{D2Q9 lattice}
  \label{fig:d2q9}
\end{figure}

@@

No slip, or zero velocity, which is commonly imposed at walls in a domain, is accomplished by simulating the particle distributions as "bouncing back" at the walls in the opposite direction from which they stream.
For example, for a particle distribution streaming in the direction of a south wall, $f_2 = f_4$, $f_5 = f_7$, and $f_6 = f_8$.
For velocity or pressure boundary conditions, the method proposed by \citet{zou1997pressure} can be used.
The particle distributions that are missing after the streaming step are solved for by assuming a bounceback of the nonequilibrium distribution in the direction normal to the boundary; e.g., for a south inlet or outlet, $f_2 - f_2^{eq} = f_4 - f_4^{eq}$.

@@

The collision operator, in the case of the continuous BE, attempts to describe the change in particle momentums and trajectories due to pairwise particle collisions (based on their respective momentums and trajectories just prior to collision) \cite{Cer90}.
In LBM, the collision operator causes particle distributions to relax toward a quasiequilibrium. % MG: is this true of all collision operators? or just BGK?
This equilibrium is determined by the macroscopic physical behavior of interest.
In the case of incompressible flow, the quasiequilibrium particle distribution, $f_i^{eq}$, is often given by:

@@

\noindent where $\omega$ is the collision frequency \cite{Bha54}.
The collision frequency can be related to macroscopic constitutive properties through Chapman-Enskog multiscale analysis.
For incompressible Newtonian flow, the relaxation time is related to the kinematic viscosity by $\nu = c_s^2(\frac{1}{\omega} - \frac{1}{2})$.
From this relationship it is clear that $\omega \in [0.0, 2.0]$, otherwise the viscosity would be negative.

Despite the utility of the BGK collision operator, it does have a few drawbacks.
For example, in low viscosity flows the BGK operator results in an over relaxation of particle distributions toward quasiequilibrium.
It is well known that when large nonequilibrium distributions exist in the LBM approximation that over relaxation can result in nonphysical oscillations that are slow to decay \cite{brownlee2007stability,dellar2003incompressible}.
To illustrate this, consider the example shown in \Fref{fig:sharp-grad}.
If $a >> 0$, then when $f_5(x,y)$ propagates to $f_5(x+1,y+1)$ during the streaming step (where $\pos = [x,y]$, $\pvel_5 = [1,1]$, and $\Delta t = 1$), $f_5(x+1,y+1)$ will be ``far'' from $f^{eq}_5(x+1,y+1)$ for the subsequent collision step.
If $\nu \rightarrow 0$ then $\omega \rightarrow 2$ and \eqref{eq:bgk} results in $f_5(x+1,y+1)$ still being ``far'' from $f^{eq}_5(x+1,y+1)$ but on the ``other side''.
If, as in \Fref{fig:sharp-grad}, the LBM macroscopic variables dictate that $f^{eq}_i$ remain mostly constant (or, at the very least, far from $f_i$) in the direction of propagation for $f_i$, then this leads to oscillations in the particle distribution, $f_i$.
An example of the nonphysical oscillations that would occur for $f_5$ in \Fref{fig:sharp-grad} can be seen in \Fref{fig:overrelaxation}.
Considering the effect oscillations of particle distributions will have on macroscopic variables and, consequently, local quasiequilibriums, positive feedback loops can occur causing the system to ``blow up'' or ``pollute'' the system enough to make the results highly nonphysical and altogether useless \cite{gorban2014enhancement}.

@@

An alternative to the BGK collision operator is the multiple-relaxation-time (MRT) collision operator.
In the LB-MRT scheme, one constructs a space based on the particle velocity, $\pvel$, moments of $\mathbf{f}$, herein referred to as the ``moment space''.
The collision is then performed in the moment space.
There a few reasons why it is advantageous to perform the collision in the moment space as opposed to the particle distribution space:
\begin{enumerate}
    \item Physical processes in fluids can be approximately described by coupling or interacting among modes; and hydrodynamic modes are directly related to, i.e. linear combinations of, mass and momentum moments \cite{lallemand2000theory}.
    \item For the D2Q9 lattice, there are nine distribution functions, $f_0, f_1, ..., f_8$, but only six variables that affect the intended hydrodynamics on a macroscopic scale, namely: $\rho$, $\mathbf{u}$, and $\mathbf{\Pi}$, where $\mathbf{\Pi}$ is the momentum flux tensor \cite{dellar2003incompressible}.
    Of the nine relaxation rates available, the three that correspond to the extra variables--often referred to as ``ghost variables'', and their associated modes as ``ghost modes''--can be tuned in order to dampen out their associated ghost modes, ensuring these modes do not dominate or cause numerical instabilities at the lattice scale.
\end{enumerate}

@@

\noindent where $\epsilon$ is related to the square of the energy $e$; $q_x$ and $q_y$ correspond to the energy fluxes in the x and y directions; and $p_{xx}$ and $p_{xy}$ correspond to the diagonal and off-diagonal component of the viscous stress tensor~\cite{lallemand2000theory}. % MG: verify this is correct
The relaxation matrix, $\relaxM$, is a diagonal matrix where each of the elements on the diagonal, $s_i \in [0, 2], i = 0, 1, ..., 8$, correspond to the relaxation rate of its associated hydrodynamic mode.
In the case when $s_0 = s_1 = ... = s_8 = \omega$ ($\omega$ is the collision frequency in the BGK sense), the MRT collision operator is equivalent to the BGK collision operator.
The relaxation parameters $s_0$, $s_3$, and $s_5$ are all set to zero as mass and momentum should be conserved.
The relaxation parameters $s_1$ and $s_7 = s_8$ are related to the bulk and shear viscosities, respectively.
The relationship for the shear viscosity is given by,
\begin{equation} \label{eq:mrt-s7}
\nu = c_s^2 \Delta t (\frac{1}{s_7} - \frac{1}{2}),
\end{equation}
\noindent which is equivalent with the relationship to the collision frequency, $\omega$, in the BGK sense (when $\Delta t = 1$).
The remaining relaxation parameters, $s_2$, $s_4$, and $s_6$, are tuned in order to dampen out and separate the ghost modes from the modes affecting hydrodynamic transport.
It is common practice, and \citet{lallemand2000theory} recommends, that these three relaxation parameters be set to slightly larger than one.

@@

Due to the discontinuous nature of \eqref{eq:bingham}, the Bingham plastic model is difficult to work with numerically.
Thus, a smooth approximation to \eqref{eq:bingham} formulated by \citet{papanastasiou1987flows} is often used as:
\begin{equation} \label{eq:papa}
\tau = \tau_y (1 - e^{-m |\dot{\gamma}|}) + \mu_p \dot{\gamma},
\end{equation}
\noindent where $m$ is the stress growth exponent.
The larger the value of $m$, the closer the approximation is to the Bingham plastic model.

@@

\chapter{Free Surface and non-Newtonian Simulation of Wellbore Cementing using the Lattice Boltzmann Method} \label{cha:lbm-cement}

@@

% what is wellbore cementing?

Cementing of oil and gas wellbores is important because if cement does not completely fill the annulus then pathways can form for fluids and gasses to flow through.
The pathways need not be large--a pathway of SOMETHING is sufficient for enough loss of pressure at the wellhead for the well to lose SOME percent production.
There are also environmental concerns as oils and gasses can migrate into ground water or the atmosphere causing environmental damage.

@@

For the sake of computational simplicity and computational efficiency, the numerical study was carried out using a two-dimensional lattice Boltzmann method on the D2Q9 lattice.
The collision operator used was the linear BGK collision operator.
For more details on the Lattice Boltzmann Method, the D2Q9 lattice, or the BGK collision operator, see \Fref{sec:LBM}.

@@

Cement slurry is best modeled using a non-Newtonian constitutive relationship.
A simple, yet accurate constitutive relationship for cement slurry is the Bingham plastic constitutive relationship.
Bingham plastics are characterized by a plastic viscosity, which is analogous to the dynamic viscosity for Newtonian fluids, and a yield stress, which is a stress threshold that, if the shear stress does not exceed, the strain-rate is zero.
The Bingham plastic constitutive model is difficult to work with numerically.
Often, a smooth approximation is used instead--the Papastasiou approximation.
Throughout this chapter the Papastasiou approximation with a stress growth exponent of $m = 10^6$ is used.
The BGK collision operator with $m = 10^6$ can be seen as a compromise between computational efficiency and accuracy for simulating Bingham plastic flows.
For more detail on Bingham plastics or the Papastasiou approximation, see \Fref{sec:bp}.

@@

The hierarchy of possible redistribution steps, though seemingly complex, ensures that mass redistribution is done in the most physically meaningful way as is possible. % We may need to hard code references to steps in here...

@@

Two important considerations not yet discussed are how boundary conditions are implemented at the interface in order to capture the interaction between the primary phase and the secondary phase, and a still more subtle consideration, how particle distributions that are missing from the streaming step are reconstructed.
The degrees of freedom, i.e. $f_i$, are not tracked for gas cells.
After the streaming step, interface cells will be missing particle distributions, $f_i$, that would have streamed from gas cells--an issue that is depicted in \Fref{fig:ps}.

>>>>

% MG: replace \~\cite with ~\cite to ensure citations stay on same line
% MG: need an overall consistency in 'flucuation' vs. 'oscillation' vs. 'noise'; use same word if you mean same thing.
% MG: same note as above for 'local equilibrium' vs. 'quasiequilibrium'
% MG: consider using collision frequency, \omega, instead of relaxation time, \tau, so that the relaxation time is not confused with shear stress

@@

\title{How the Lattice Boltzmann Method changed my life forever!}
% The optional argument is the version of the title that will appear in Acrobat Reader's Document Info dialog box.
\author{Matthew Grasinger}
\degree{BS in Civil and Environmental Engineering, University of Pittsburgh, Pittsburgh, PA 2013}
\date{July 20, 2016}
\keywords{lattice Boltzmann method, non-Newtonian flow, free surface flow, primary cementing}
\subject{Draft}
\school{Swanson School of Engineering}
\degreesought{Master of Science}

@@

Despite the advantages of LBM, there are considerations that must be made, as with all computational methods, in regards to the accuracy and numerical stability of the solution.
Issues with accuracy and numerical stability are especially prevalent in non-Newtonian flows because of the nonlinear constitutive relationship.
\Fref{chap:lbm-stability} is a numerical investigation of the accuracy, stability, and computational efficiency of different LBM models in simulating non-Newtonian flows.

Once the ground-work has been laid for simulating non-Newtonian flows using LBM, the algorithm was extended in \Fref{chap:lbm-cement} to incorporate simulation of free-surface flows.
The extended LBM model was used to study primary cementing of a dry annulus, i.e. an annulus that is not filled with drilling mud.
More specifically, the study involved parameterizing different cement slurry flows and investigating how well each slurry flow filled different wellbore geometries.
The study is followed by conclusions and a discussion of future work.

@@

\chapter{Numerical Investigation of the Accuracy, Stability, and Efficiency of Lattice Boltzmann Models in Simulating Non-Newtonian Flows} \label{chap:lbm-stability}

@@

The dynamic viscosity of a fluid is a measure of its resistance to shear deformation.
For many fluids, at a constant temperature, the dynamic viscosity can be considered as constant.
These fluids are known as Newtonian fluids.
Non-Newtonian fluids have an apparent dynamic viscosity, or apparent resistance to shear stress, that is variable--even at a constant temperature--and is often a function of strain-rate.
There are a number of fluids in science and engineering applications that can be classified as non-Newtonian; pastes, slurries, molten plastics, polymer solutions, dyes, varnishes, suspensions, and some biomedical liquids such as blood all behave in a non-Newtonian manner~\cite{bohme1987non}.
Of all of the different non-Newtonian behaviors that exist, there are two models under which much of the behaviors may be idealized:
\begin{enumerate}
	\item \emph{Yield stress fluids}. Yield stress fluids, also known as Bingham plastics, do not flow until a threshold value of stress, referred to as its yield stress, is exceeded.
	Some examples of yield stress fluids are drilling muds, fresh cement slurries and concretes, and melted metal alloys~\cite{ginzburg2002free}.
	Understanding the flow of cement slurries and concretes is important because they are used in the cementing of oil and gas wellbores, and in construction.
	If the cement job for an oil and gas wellbore is not completed successfully then gases and fluids can leak through the wellbore to contaminate ground water or the atmosphere~\cite{grasinger2015simulation}. %MG: is it tacky to cite myself here?
	The goal of understanding cement slurry and concrete flows during construction is to understand how construction forms can be completely filled as voids can lead to structural deficiencies.
	Yield stress flow is relevant in many other disciplines and applications because of the many substances that exhibit yield stress behavior, e.g. pastes, paints, muds, molten plastics and metals, and in some cases, blood~\cite{wang2011lattice}. %MG: citation is for blood as non-Newtonian
	As a consequence, simulating yield stress fluids can help engineers develop better pastes and paints, understand the flow and deformation of mud and clay in geotechnical engineering, understand blood circulation, and better manufacture metal and plastic parts.

@@

Analytical solutions rarely exist for even the simplest non-Newtonian flows because of the complexity that a nonlinear constitutive relationship entails.
It is generally more practical to approximate non-Newtonian flows using numerical methods.
However, the nonlinear constitutive equation--typically of the form \begin{equation} \label{eq:non-newtonian}
\tau = \mu_{app}(\dot{\gamma}) \dot{\gamma},
\end{equation}\noindent where the apparent viscosity, $\mu_{app}$ is a function of the strain rate--results in certain challenges for numerical methods as well.
%Constitutive form for generalized Newtonian fluids, $\tau = \mu(\dot{\gamma}) \dot{\gamma}$, where the apparent viscosity, $\mu$ is a function of the strain rate
Determining the apparent viscosity and shear rate of a flow will often require an iterative solution, a general Picard-type algorithm is:

@@

One such consideration associated with this view is the potential for numerical inaccuracies and instabilities ~\cite{sterling1993stability,sterling1996stability,bawazeer2013stability,lallemand2000theory}. %MG: are all relevant sources cited?

@@

Much work has been done (not specific to non-Newtonian flows) to investigate and develop ways to enhance the stability of lattice Boltzmann methods.
\citet{sterling1993stability,sterling1996stability} investigated the stability of LBM with the linear Bhatnagar--Gross--Krook (BGK) collision operator on 7-velocity and 9-velocity 2D lattices and a 15-velocity cubic lattice.
The investigation was performed by linearizing all nonlinear terms about global equilibrium values and applying a von Neumann stability analysis on the resulting linearized equations.
The spatial dependence of stability was analyzed by taking the Fourier transform of the linearized equations.
\citeauthor{sterling1993stability} concluded that the linear stability of LBM models depends on the mass distribution parameters, the mean velocity, the relaxation time, and the wave number.
~\citet{worthing1997stability} extended the stability analysis performed by~\citeauthor{sterling1993stability} and discovered both physical and nonphysical instabilities; and that setting the BGK collision frequency, $\omega = 1$ provides the optimal accuracy in time.

@@

To improve upon the stability of LBM models,~\citet{d1994generalized} developed a multiple-relaxation-time (MRT) collision operator, that takes place in moment space and allows each moment to relax at a different rate.
\citet{lallemand2000theory} used von Neumann stability analysis to investigate the stability of the newly constructed LBM-MRT model and concluded LBM with the MRT collision operator was more stable, but with increased computational expense, than what is still often used, the BGK collision operator.
Note that although this increased computational expense was decided to be not too significant for Newtonian flows ($\approx$ 10-20\%~\cite{lallemand2000theory}), the issue may be magnified for non-Newtonian flows because an iterative solution for the constitutive equation can require that certain expensive computations be performed at each iteration (the strain rate tensor is determined from the nonequilibrium particle distribution which must be mapped into moment space when using the MRT collision operator).
Both the BGK~\cite{wang2011lattice,ashrafizaadeh2009comparison,wang2015localized,tang2011bingham} and MRT~\cite{chen2014simulations,fallah2012multiple,chai2011multiple} collision operators have been used successfully to simulate non-Newtonian flows.
\citeauthor{chen2014simulations} concluded that the MRT collision operator was more stable for Bingham plastic flow and allowed the use of a more accurate approximation to the Bingham plastic constitutive relationship.
However, what remains unclear is:
\begin{itemize}
    \item What is the increased cost associated with the MRT collision operator when applied to non-Newtonian flows?
    \item Under what conditions, e.g. material parameters, physical problem, etc., for Bingham plastic fluid flows is the MRT collision operator necessary to maintain stability and/or accuracy?
    \item Under what conditions, e.g. material parameters, physical problem, etc, for power-law fluid flows is the MRT collision operator necessary to maintain stability and/or accuracy?
    \item What are additional strategies for increasing stability and accuracy, and what are their associated computational costs?
\end{itemize}
In regards to the last question, much work has been done recently to enhance stability of LBM models beyond the MRT collision operator.
\citet{brownlee2006stabilization,brownlee2007stability,brownlee2008nonequilibrium,packwood2009entropy,gorban2014enhancement} have all developed and tested means for introducing artificial dissipation in order to dampen out high frequency, nonphysical oscillations--oscillations that often lead to numerical instabilities if left unchecked.
Stability enhancement through artificial dissipation and entropic filtering has shown a lot of promise, but to the authors' knowledge has not been tested for use in non-Newtonian flows.

@@

\subsection{The Boltzmann Equation}
The Boltzmann equation (BE) can be thought of as a conservation of particle distributions.
The BE is given as $\frac{\partial f}{\partial t} + \pvel \cdot \grad f = \colop$, where $f = f(\pos, \pvel, t)$ is the particle velocity distribution function, $\pos$ is the spatial position vector, $\pvel$ is the particle velocity, and $\colop$ is the collision operator.
The lattice part of LBM refers to the way in which the BE is discretized.
The lattice discretizes the spatial domain with nodes that are connected to their neighbors through discrete lattice velocity vectors.
The velocity vectors act as pathways for particle distributions to travel along.
Each time step in LBM consists of two distinct actions:

\begin{itemize}
\item Streaming: particle distributions propagate to their neighbors along the lattice velocity vectors.
The particles can only move along the vectors in their specified direction and can only move at a specific speed.
\item Collision: particle distributions meet at a node and ``collide''.
In LBM, collisions are not simulated in a realistic sense, meaning that each individual particle does not exist and glance off of, or interact with, one another.
Instead the collision operator is formulated in such a way that particle distributions are relaxed toward equilibrium.
What defines equilibrium depends on the mechanics of interest to be modeled.
\end{itemize}

@@

\begin{figure}
  \centering
  \includegraphics[width=10cm]{figs/d2q9}
  \caption{D2Q9 lattice}
  \label{fig:d2q9}
\end{figure}

@@

No slip, or zero velocity, which is commonly imposed at walls in a domain, is accomplished by simulating the particle distributions as "bouncing back" at the walls in the opposite direction from which they stream.
For example, for a particle distribution streaming in the direction of a south wall, $f_2 = f_4$, $f_5 = f_7$, and $f_6 = f_8$.
For velocity or pressure boundary conditions, the method proposed by~\citet{zou1997pressure} can be used.
The particle distributions that are missing after the streaming step are solved for by assuming a bounceback of the nonequilibrium distribution in the direction normal to the boundary; e.g., for a south inlet or outlet, $f_2 - f_2^{eq} = f_4 - f_4^{eq}$.

@@

The collision operator, in the case of the continuous BE, attempts to describe the change in particle momentums and trajectories due to pairwise particle collisions (based on their respective momentums and trajectories just prior to collision)~\cite{Cer90}.
In LBM, the collision operator causes particle distributions to relax toward a quasiequilibrium. % MG: is this true of all collision operators? or just BGK?
This equilibrium is determined by the macroscopic physical behavior of interest.
In the case of incompressible flow, the quasiequilibrium particle distribution, $f_i^{eq}$, is often given by:

@@

\noindent where $\omega$ is the collision frequency~\cite{Bha54}.
The collision frequency can be related to macroscopic constitutive properties through Chapman-Enskog multiscale analysis.
For incompressible Newtonian flow, the relaxation time is related to the kinematic viscosity by $\nu = c_s^2(\frac{1}{\omega} - \frac{1}{2})$.
From this relationship it is clear that $\omega \in [0.0, 2.0]$, otherwise the viscosity would be negative.

Despite the utility of the BGK collision operator, it does have a few drawbacks.
For example, in low viscosity flows the BGK operator results in an over relaxation of particle distributions toward quasiequilibrium.
It is well known that when large nonequilibrium distributions exist in the LBM approximation that over relaxation can result in nonphysical oscillations that are slow to decay~\cite{brownlee2007stability,dellar2003incompressible}.
To illustrate this, consider the example shown in \Fref{fig:sharp-grad}.
If $a >> 0$, then when $f_5(x,y)$ propagates to $f_5(x+1,y+1)$ during the streaming step (where $\pos = [x,y]$, $\pvel_5 = [1,1]$, and $\Delta t = 1$), $f_5(x+1,y+1)$ will be ``far'' from $f^{eq}_5(x+1,y+1)$ for the subsequent collision step.
If $\nu \rightarrow 0$ then $\omega \rightarrow 2$ and \eqref{eq:bgk} results in $f_5(x+1,y+1)$ still being ``far'' from $f^{eq}_5(x+1,y+1)$ but on the ``other side''.
If, as in \Fref{fig:sharp-grad}, the LBM macroscopic variables dictate that $f^{eq}_i$ remain mostly constant (or, at the very least, far from $f_i$) in the direction of propagation for $f_i$, then this leads to oscillations in the particle distribution, $f_i$.
An example of the nonphysical oscillations that would occur for $f_5$ in \Fref{fig:sharp-grad} can be seen in \Fref{fig:overrelaxation}.
Considering the effect oscillations of particle distributions will have on macroscopic variables and, consequently, local quasiequilibriums, positive feedback loops can occur causing the system to ``blow up'' or ``pollute'' the system enough to make the results highly nonphysical and altogether useless~\cite{gorban2014enhancement}.

@@

An alternative to the BGK collision operator is the multiple-relaxation-time (MRT) collision operator.
In the LB-MRT scheme, one constructs a space based on the particle velocity, $\pvel$, moments of $\mathbf{f}$, herein referred to as the ``moment space''.
The collision is then performed in the moment space.
There a few reasons why it is advantageous to perform the collision in the moment space as opposed to the particle distribution space:
\begin{enumerate}
    \item Physical processes in fluids can be approximately described by coupling or interacting among modes; and hydrodynamic modes are directly related to, i.e. linear combinations of, mass and momentum moments~\cite{lallemand2000theory}.
    \item For the D2Q9 lattice, there are nine distribution functions, $f_0, f_1, ..., f_8$, but only six variables that affect the intended hydrodynamics on a macroscopic scale, namely: $\rho$, $\mathbf{u}$, and $\mathbf{\Pi}$, where $\mathbf{\Pi}$ is the momentum flux tensor~\cite{dellar2003incompressible}.
    Of the nine relaxation rates available, the three that correspond to the extra variables--often referred to as ``ghost variables'', and their associated modes as ``ghost modes''--can be tuned in order to dampen out their associated ghost modes, ensuring these modes do not dominate or cause numerical instabilities at the lattice scale.
\end{enumerate}

@@

\noindent where $\epsilon$ is related to the square of the energy $e$; $q_x$ and $q_y$ correspond to the energy fluxes in the x and y directions; and $p_{xx}$ and $p_{xy}$ correspond to the diagonal and off-diagonal component of the viscous stress tensor~\cite{lallemand2000theory}. % MG: verify this is correct
The relaxation matrix, $\relaxM$, is a diagonal matrix where each of the elements on the diagonal, $s_i \in [0, 2], i = 0, 1, ..., 8$, correspond to the relaxation rate of its associated hydrodynamic mode.
In the case when $s_0 = s_1 = ... = s_8 = \omega$ ($\omega$ is the collision frequency in the BGK sense), the MRT collision operator is equivalent to the BGK collision operator.
The relaxation parameters $s_0$, $s_3$, and $s_5$ are all set to zero as mass and momentum should be conserved.
The relaxation parameters $s_1$ and $s_7 = s_8$ are related to the bulk and shear viscosities, respectively.
The relationship for the shear viscosity is given by,
\begin{equation} \label{eq:mrt-s7}
\nu = c_s^2 \Delta t (\frac{1}{s_7} - \frac{1}{2}),
\end{equation}
\noindent which is equivalent with the relationship to the collision frequency, $\omega$, in the BGK sense (when $\Delta t = 1$).
The remaining relaxation parameters, $s_2$, $s_4$, and $s_6$, are tuned in order to dampen out and separate the ghost modes from the modes affecting hydrodynamic transport.
It is common practice, and~\citet{lallemand2000theory} recommends, that these three relaxation parameters be set to slightly larger than one.

@@

Due to the discontinuous nature of \eqref{eq:bingham}, the Bingham plastic model is difficult to work with numerically.
Thus, a smooth approximation to \eqref{eq:bingham} formulated by~\citet{papanastasiou1987flows} is often used as:
\begin{equation} \label{eq:papa}
\tau = \tau_y (1 - e^{-m |\dot{\gamma}|}) + \mu_p \dot{\gamma},
\end{equation}
\noindent where $m$ is the stress growth exponent.
The larger the value of $m$, the closer the approximation is to the Bingham plastic model.

@@

\chapter{Free Surface and non-Newtonian Simulation of Wellbore Cementing using the Lattice Boltzmann Method} \label{chap:lbm-cement}

@@

% what is primary cementing?

@@

For the sake of computational simplicity and computational efficiency, the numerical study considered a two-dimensional section of the wellbore and was simulated using the D2Q9 lattice.
In addition, the collision operator used was the BGK collision operator.
For more details on the Lattice Boltzmann Method, the D2Q9 lattice, or the BGK collision operator, see \Fref{sec:LBM}.

@@

Cement slurry is best modeled using a non-Newtonian constitutive relationship.
A simple, yet accurate constitutive relationship for cement slurry is the Bingham plastic constitutive relationship.
Bingham plastics are characterized by a plastic viscosity, which is analogous to the dynamic viscosity for Newtonian fluids, and a yield stress, which is a stress threshold that, if the shear stress does not exceed, the strain-rate is zero.
The Bingham plastic constitutive model is difficult to work with numerically.
Often, a smooth approximation is used instead--the Papastasiou approximation.
Throughout this chapter, the Papastasiou approximation with a stress growth exponent of $m = 10^6$ is used.
The BGK collision operator with $m = 10^6$ can be seen as a compromise between computational efficiency and accuracy for simulating Bingham plastic flows.
For more detail on Bingham plastics or the Papastasiou approximation, see \Fref{sec:bp}.

@@

The hierarchy of possible redistribution steps ensures that mass redistribution is done in the most physically meaningful way as is possible. % We may need to hard code references to steps in here...

@@

Two important considerations not yet discussed are how boundary conditions are implemented at the interface in order to capture the interaction between the primary phase and the secondary phase, and a still more subtle consideration, how particle distributions that are missing from the streaming step are reconstructed.
The degrees of freedom, $f_i$, are not tracked for gas cells.
After the streaming step, interface cells will be missing particle distributions that would have streamed from gas cells--an issue that is depicted in \Fref{fig:ps}.
